Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2017-12-20 article.tex
Los cambios de hoy

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradient Orientations of the Signal Plot method is explained. Section~\ref{Pipeline} describes the preprocessing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the Speller Matrix Letter Identification procedure including the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.
%This kind of EEG events are described in~\citep{Hartman2005} and~\citep{Ben-Simon2013}.

@@

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, t \in \{0, 1, \dots, t_{max}\},c \in \{1,2,\dots,Ch\}
\label{eq:standarizedaverages}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. In addition, $ x(t,c) $ is the point-to-point averaged EEG matrix for time $t$ and for channel $c$. The parameter  $Ch$ is the number of available EEG channels. Lastly, $$\bar{x}(c) =\frac{1}{t_{max}}\sum_{t=1}^{t_{max}}x(t,c)$$ and $$ \sigma(c) = \frac{1}{t_{max}}\sum_{t=1}^{t_{max}}(x(t,c)-\bar{x}(c))^2 $$ are the mean and standard deviation of $x(t,c), t \in 1,\dots, t_{max}$, for each channel $c$.

@@

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \cdot t; z_2 = \tilde{x}(t,c) + z(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

\noindent where $ (z_1,z_2) \in \mathcal{N} \times \mathcal{N}$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image,  $t \in \{0, 1, \dots, t_{max}\}$ and $c \in \{1,2,\dots,Ch\}$.  The function $z(c)$, $c \in \{1,2,\dots,Ch\}$ is the location on the image where the signal's zero value has to be located in order to fit the entire signal within the image:

\begin{equation}
z(c) = \left \lfloor{ \frac{\max_{0 \leq t\leq t_{max}} \tilde{x}(t,c)  - \min_{0 \leq t\leq t_{max}} \tilde{x}(t,c) }{2} }\right \rfloor -   \left \lfloor{ \frac{\max_{0 \leq t\leq t_{max}} \tilde{x}(t,c)  + \min_{0 \leq t\leq t_{max}} \tilde{x}(t,c)}{ 2} }\right \rfloor
\label{eq:zerolevel}
\end{equation}

@@

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t \in \{0, 1, \dots, t_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to time and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c), t \in \{0, 1, \dots, t_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

\item \label{paso4}Plot the signal $x(t,c)$,  $t \in \{0, 1, \dots, t_{max}\}$, $c \in \{1,\dots,Ch\}$,  according Section~\ref{Plot}.

@@

The patch size is $S_P = 3s.4 \times 3s.4$ pixels, where $s$ is the scale of the local patch and it is an input parameter of the algorithm. The P300 event can have a span of $400$ ms and its amplitude can reach $ 10 \mu V $~\citep{Rao2013}.  Hence it is necessary to utilize a size patch $S_P$ that could capture an entire transient event. With this purpose in consideration, the $s$ value election is essential.

@@

Lastly, the number of channels $Ch$ is equal to $8$ for both datasets, and the number of intensification sequences $k_a$ is statically assigned to $10$.  The parameter $k$ used in the Near Neighbor $NN_T(d^l_u), \;l\in\{row,{col}\}$ is set to $k=7$.  In addition, the norm used on  Equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol} was the cosine norm, and descriptors were normalized to $ \left[ -1, 1 \right] $.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

@@

A method to detect transient P300 components from EEG signals based on their waveform characterization in time-space was presented.  Additionally its validity was verified against a public dataset of ALS patients and an own dataset of healthy subjects.

This method can be both, based on predefined templates or adaptive by including a training or calibration step to identify the templates. The adaptive behavior of the algorithm make it well suited when the shape of the pattern elicited by the P300 response does not conform to the predicted structure.

The method works on a channel by channel basis.  Should a calibration step be included, it can be used to identify the best performing channel, and may drastically reduce the number of required EEG electrodes, leading to the use of more ergonomic capturing device.

We verified that the shape of the wave is more stable in occipital channels, where the performance at identifying letters  is higher.  Besides novel applications of BCI, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.  It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.  It may be worthy of further consideration to evaluate if there is any correlation between the lower performance obtained for some subjects and the characterization of their ALS stage.  Althought this, we did not find such an evidence in terms of the shape.  The best response was obtained by an ALS subject.  Moreover, this method can be used as an alternate \textit{BCI predictor} \citep{Clerc}, i.e. to detect BCI illiteracy or to predict the achievable performance of a given method.  In the original work~\citep{Riccio2013} where the dataset was published, a link between a task prediction test and the performance of the p300 was found.  At the same time, by analyzing the generated descriptors, which map in a very detailed and synthetic structure the shape information contained within the patch, a metric about the consistency of the \textit{shape} of the generated P300 response could also be derived.

We also found that the stability of the P300 in terms of shape is crucial: Synchronization averaging, increase inter-stimulus interval, different montages, spatial filters can all of them affect the generated shape, and it is advisable to study which methods or combinations are best in order to stabilize the shape of the p300.

Clearly, the best characteristic of this method is that closer collaboration with physicians can be fostered.
We believe that the expanding and the understanding of this tool in order to automatically classify those patterns in EEG that are specifically identified by their shapes (e.g. K-Complex, Vertex Waves, Positive Occipital Sharp Transient \citep{Hartman2005}) is a prospect future work to be considered. We are currently working in unpublished material analyzing KComplex that may also provide  assistance to physician or electroencephalographers to help them locate these EEG patterns particularly in long recording periods, frequent in sleep research.  Additionally, it can be used as a tool for artifact removal (which is performed on many occasions by visually inspecting the signal).  This is due to the fact that the descriptors are directly based on how the signals behave in shape domain (i.e. \textit{how actually they looked like}).  In line with these applications,  it can be used to build a database~\citep{Chavarriaga2017} of descriptors and improve atlases~\citep{Hartman2005}.

In terms of the tool itself, the SIFT method is invariant to the scale, so in order to allow our method to be also invariant to the amplitude of the signal we had to modify the geometry of the patch to remove the squared constrain.

Finally, waveform analysis and the methods that can be used to analyze them is an area where active research can benefit greatly both, physicians who need to automate some of the tools and practices and Brain Computer Interfaces or automatic EEG processing by using the extensive knowledge body available for clinical practices~\citep{Cole2017}

>>>>

This article unfolds as follows: Section~\ref{Feature} is dedicated to explain the Feature Extraction method based on Histogram of Gradient Orientations of the Signal Plot: Section~\ref{Pipeline} shows the preprocessing pipeline,  Section~\ref{Plot}  describes the image generation of the signal plot, Section~\ref{SIFT}  presents the feature extraction procedure while  Section~\ref{Classification}  introduces the whole  Speller Matrix Letter Identification procedure including the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}. In Section~\ref{Protocol}, the experimental protocol is expounded. Section~\ref{Results} shows the results of applying the proposal technique and a discussion is presented. 
In the final Section~\ref{conclusion}  we expose our remarks, conclusions and future work.
%This kind of EEG events are described in~\citep{Hartman2005} and~\citep{Ben-Simon2013}.

@@

\begin{equation}
\tilde{x}(n,c) = \left \lfloor{ \gamma \cdot \frac{( x(n,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, n \in \{ 1, \dots, n_{max}\},c \in \{1,2,\dots,Ch\}
\label{eq:standarizedaverages}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. In addition, $ x(n,c) $ is the point-to-point averaged EEG matrix for the sample point $n$ and for channel $c$, $n_{max}=F_s.t_{max}$ and $F_s$ is the sampling frequency. The parameter  $Ch$ is the number of available EEG channels. Lastly, $$\bar{x}(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x(n,c)$$ and $$ \sigma(c) = (\frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}}(x(n,c)-\bar{x}(c))^2 )^{\frac{1}{2}}$$ are the mean and standard deviation of $x(n,c), n \in 1,\dots,n_{max}$, for each channel $c$.

@@

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \cdot n; \; z_2 = \tilde{x}(n,c) + z(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

\noindent where $ (z_1,z_2) \in \mathbb{N} \times \mathbb{N}$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image,  $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$.  The function $z(c)$, $c \in \{1,2,\dots,Ch\}$ is the location on the image where the signal's zero value has to be located in order to fit the entire signal within the image:

\begin{equation}
z(c) = \left \lfloor{ \frac{\max_{1 \leq n\leq n_{max}} \tilde{x}(n,c)  - \min_{1 \leq n\leq n_{max}} \tilde{x}(n,c) }{2} }\right \rfloor -   \left \lfloor{ \frac{\max_{1 \leq n\leq n_{max}} \tilde{x}(t,c)  + \min_{1 \leq n\leq n_{max}} \tilde{x}(n,c)}{ 2} }\right \rfloor
\label{eq:zerolevel}
\end{equation}

@@

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(n,c),\dots,S_{k_a}(n,c)$, where the variables $n \in \{1, \dots, n_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to time and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x(n,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(n,c),n \in \{1, \dots, n_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

\item \label{paso4}Plot the signal $x(n,c)$,  $n \in \{1, \dots, n_{max}\}$, $c \in \{1,\dots,Ch\}$,  according Section~\ref{Plot}.

@@

The patch size is $S_P = 4.3s \times 4.3s$ pixels, where $s$ is the scale of the local patch and it is an input parameter of the algorithm. The P300 event can have a span of $400$ ms and its amplitude can reach $ 10 \mu V $~\citep{Rao2013}.  Hence it is necessary to utilize a size patch $S_P$ that could capture an entire transient event. With this purpose in consideration, the $s$ value election is essential.

@@

Lastly, the number of channels $Ch$ is equal to $8$ for both datasets, and the number of intensification sequences $k_a$ is statically assigned to $10$.  The parameter $k$ used in the Near Neighbor $NN_T(d^l_u), \;l\in\{row,{col}\}$ is set to $k=7$, following the suggestion of the article~\citep{Boiman2008}.  In addition, the norm used on  Equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol} was the cosine norm, and descriptors were normalized to $ \left[ -1, 1 \right] $.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{conclusion}

@@

Among other applications of BCI analysis, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.

In this work, a method to detect transient P300 components from EEG signals based on their waveform characterization in digital time-space, is presented.  Additionally, its validity is evaluated using a public dataset of ALS patients and an own dataset of healthy subjects. The objective of this article is to explore new techniques of P300 automatic interpretation.  
We know that we have to keep working but the results are promising.

The method works on a channel by channel basis; in this way the best performing channel can be identified and used it to reduce the number of required EEG electrodes, leading to the development of more ergonomic capturing device.

We observed that the shape of the wave is more stable in occipital channels, where the performance for identifying letters is higher.   % It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.  
It is important to evaluate if there is any correlation between the lower performance obtained for some subjects and the characterization of their ALS stage.  Although,  we did not find any evidence in terms of the shape digital signal and the best response was obtained for an ALS subject.  Moreover, this method can be used as an alternate BCI predictor.

By analyzing the generated descriptors, a metric about the shape of the P300 response can also be derived.

By means of the empirical experiments, we concluded that the stability of the P300 in terms of shape is crucial: synchronization averaging,  inter-stimulus interval,  montages, the signal to noise ratio and spatial filters can all of them affect the stability of the shape of the P300 ERP.

In our opinion, the best application of this approach is that a closer collaboration with physicians can be fostered, because our method intent to imitate human visual observation.
Automatically classify those patterns in EEG that are specifically identified by their shapes (e.g. K-Complex, Vertex Waves, Positive Occipital Sharp Transient~\citep{Hartman2005}) is a prospect future work to be considered. We are currently working in unpublished material analyzing KComplex that also provide  assistance to physician to locate these EEG patterns, specially in long recording periods, frequent in sleep research.  
Additionally, it can be used for artifact removal which is performed on many occasions by visually inspecting the signal.  This is due to the fact that the descriptors are directly based on the signals behavior in shape domain.  In line with these applications,  it can be used to build a database~\citep{Chavarriaga2017} of descriptors and improve atlases~\citep{Hartman2005}.

We also want to solve the problem of signal stabilization by applying different shifts on the averaging procedure and another methods for obtaining a correct synchronization.

%Finally, waveform analysis and the methods that can be used to analyze them is an area where active research can benefit greatly both, physicians who need to automate some of the tools and practices and Brain Computer Interfaces or automatic EEG processing by using the extensive knowledge body available for clinical practices~\citep{Cole2017}
